---
title: "Applied regression modeling in R"

authors:   
  - name: "Guy-Alain Schnidrig (VPHI)"
  - name: "Filipe Miguel Maximiano Alves de Sousa (VPHI)"
  - name: "Eleftheria Michalopoulou (ISPM)"
  - name: "Beatriz Vidondo (VPHI)"
  - name: "Ben Spycher (ISPM)"

toc: true
toc-depth: 10
toc-location: left
number-sections: false
editor: source
format: 
  html:
    self-contained: true
    code-fold: true
    theme: flatly
---

### Information

Exercises (solution R-code provided) for [Linear and Logistic Regression Modelling in R](https://zuw.me/kurse/dt.php?kid=4476) course of the [Public Health Sciences Course Program](https://www.medizin.unibe.ch/studies/study_programs/phs_course_program) at the [University of Bern](https://www.unibe.ch/).

To follow along with the course, we recommend that you open a new R-Script (or Quarto/Markdown). 
Before starting the exercises, you should run the provided setup chunk in at least once.


```{r setup, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE}
# Set language
Sys.setenv(LANG = "en")

# Clear memory
rm(list=ls())
gc()

# Load libraries
library_names <- c("survival", "knitr", "nnet", "MASS", "TH.data", "datasets")

install_or_load_pack <- function(pack){
  create.pkg <- pack[!(pack %in% installed.packages()[, "Package"])]
  if (length(create.pkg)) install.packages(create.pkg, dependencies = TRUE, repos="https://cloud.r-project.org")
  sapply(pack, require, character.only = TRUE)
}

install_or_load_pack(library_names)
rm(library_names)

```


<!-- #.....................................................................# -->
<!-- #.....................................................................# -->


### Load data

For this series of exercises we will be using the following data: 

* **For exercise 1:** data *Melanoma* from package `MASS` (time-to-event data on patients diagnosed with malignant melanoma in Denmark)

* **For exercise 2:** data *mammoexp* from package `TH.data` (data from a questionnaire on the benefits of mammography)

* **For exercise 3:** data *infert* from package `datasets` (matched case-control study that investigated the role of abortions in the aetiology of secondary infertility)

```{r}
#| code-fold: false

# Exercise 1
data(Melanoma, package = "MASS")

# Exercise 2
data(mammoexp, package = "TH.data")

# Exercise 3
data(infert, package = "datasets")


```


<!-- #.....................................................................# -->
<!-- #.....................................................................# -->


## Exercise 1

Using data from 205 patients diagnosed with malignant melanoma in Denmark, we would like to explore the odds of dying (versus staying alive) in relationship to the presence of tumor ulceration. Ulceration is a breakdown of the skin on top of the melanoma and is an independent predictor of melanoma survival according to the literature. 

For our purposes we assume that patients' status was measured at a fixed period after melanoma diagnosis, such that probabilities can be interpreted as risks.


<!-- #.....................................................................# -->

#### (a) Manual calculation of odds ratios
Cross-tabulate the variables `status` and `ulcer` to manually calculate and interpret 

i) the odds ratio of dying from melanoma versus staying alive and 
ii) the odds ratio of dying from other causes versus staying alive in patients with tumor ulceration compared to patients without ulceration. 

```{r}

Melanoma$status <- factor(Melanoma$status, levels = c(1, 2, 3), labels = c(
  "death from melanoma", "alive", "death from other causes"))

Melanoma$ulcer <- factor(Melanoma$ulcer, levels = c(0, 1), 
                         labels = c("absent", "present"))


tab_statUlcer <- table(Melanoma$status, Melanoma$ulcer)

mosaicplot(tab_statUlcer, main = "Melanoma", xlab = "status",
           ylab = "ulceration")


tab_statUlcer_pt <- rbind(tab_statUlcer, colSums(tab_statUlcer))
tab_statUlcer_pt <- cbind(tab_statUlcer_pt, rowSums(tab_statUlcer_pt))


row.names(tab_statUlcer_pt) <- c(row.names(tab_statUlcer), "total")

colnames(tab_statUlcer_pt) <- c(colnames(tab_statUlcer), "total")

tab_statUlcer_pt


# OR_12 = ((41/90) / (42/90)) / ((16/115) / (92/115)) = (41/42) / (16/92)
OR_12 <- (41 * 92) / (42 * 16) 
OR_12

# OR_32 = ((7/90) / (42/90)) / ((7/115) / (92/115)) = (7/42) / (7/92)
OR_32 <- (7 * 92) / (42 * 7) 
OR_32

```


<!-- #.....................................................................# -->

#### (b) Univariable multinomial model
Fit a multinomial model with status as dependent variable and ulcer as explanatory variable. Which estimated coefficients show the odds ratios calculated in (a)? Calculate their 95% confidence intervals. Is there evidence for an association between tumor ulceration and the different odds of death?

**Hints:** to set the reference levels of a factor, use function `relevel()`. To calculate confidence intervals for model estimated coefficients, use function `confint()`.
```{r}

# Set the reference levels of the variables "status" and 
# "ulcer":
Melanoma$status <- relevel(Melanoma$status, ref = "alive")
Melanoma$ulcer <- relevel(Melanoma$ulcer, ref = "absent")


multinom_melanom <- multinom(status ~ ulcer, data = Melanoma)

summary(multinom_melanom)


confint(multinom_melanom, level = 0.95)

```


<!-- #.....................................................................# -->

#### (c) Multivariable multinomial model
Assess again the association of tumor ulceration with the different odds of dying versus staying alive after adjusting the multinomial model for patients' age and sex. Check how the estimated odds ratios change and calculate their confidence intervals.

```{r, warning = FALSE, message = FALSE, include = FALSE}

multinom_melanom2 <- multinom(status ~ ulcer + age + sex, data = Melanoma)

```

```{r}

summary(multinom_melanom2)

confint(multinom_melanom2, level = 0.95)

```


<!-- #.....................................................................# -->

#### (d) Odds estimated from the multivariable multinomial model
What is the odds of dying from melanoma for a 75-year old man without tumor ulceration?

**Hint:** the reference level of variable sex is 0 = female.
```{r}

exp(-2.703762 + 0.01474460 * 75 + 0.5749964) 

```


<!-- #.....................................................................# -->

#### (e) Probabilities estimated from the multivariable multinomial model
Calculate the probability of dying from melanoma for a 50-year old woman with tumor ulceration.

**Hint:** The model estimates the following two log odds:

$$log \biggl(\frac{P[status = \textit{death from melanoma} \ | \ ulcer = present]}{P[status = alive \ | \ ulcer = present]} \biggr)$$
$$log \biggl(\frac{P[status = \textit{death from other causes} \ | \ ulcer = present]}{P[status = alive \ | \ ulcer = present]} \biggr)$$
and we also know that
$$P[status = \textit{death from melanoma} \ | \ ulcer = present] + P[status = \textit{death from other causes} \ | \ ulcer = present] + P[status = alive \ | \ ulcer = present] = 1$$
Calculate the two odds for a 50-year old woman with tumor ulceration and then solve the system of the three equations above to estimate the probability in question.

```{r}

# The odds of dying from melanoma versus staying alive for a 50-year old woman
# with tumor ulceration is:
odds_deathMel <- exp(-2.703762 + 1.6152588 + 0.01474460 * 50) 
odds_deathMel # 0.7037915

# This odds is equal to the probability of dying from melanoma over the 
# probability of staying alive for a 50-year old woman with tumor ulceration:

# P[status = "death from melanoma" | ulcer = "present"] / P[status = "alive" | ulcer = "present"] = 0.7037915

# meaning that:
# P[status = "death from melanoma" | ulcer = "present"] = 0.7037915 * P[status = "alive" | ulcer = "present"]


# The odds of dying from other causes versus staying alive for a 50-year old 
# woman with tumor ulceration is:
odds_deathOther <- exp(-6.419789 + 0.4821069 + 0.06459613 * 50) 
odds_deathOther # 0.06667831

# This odds is equal to the probability of dying from from other causes over the 
# probability of staying alive for a 50-year old woman with tumor ulceration.

# P[status = "death from other causes" | ulcer = "present"] / P[status = "alive" | ulcer = "present"] = 0.06667831

# meaning that:
# P[status = "death from other causes" | ulcer = "present"] = 0.06667831 * P[status = "alive" | ulcer = "present"]


# We also kniw that: 
# P[status = "death from melanoma" | ulcer = "present"] + 
#   P[status = "death from other causes" | ulcer = "present"] +
#   P[status = "alive" | ulcer = "present"] = 1

# meaning that:
# P[status = "alive" | ulcer = "present"] = 1 - P[status = "death from melanoma" | ulcer = "present"] -
#   P[status = "death from other causes" | ulcer = "present"]
# P[status = "alive" | ulcer = "present"] = 1 - 0.06667831 * P[status = "alive" | ulcer = "present"] -
#   0.7037915 * P[status = "alive" | ulcer = "present"]
# P[status = "alive" | ulcer = "present"] = 1 - 0.06667831 * P[status = "alive" | ulcer = "present"] -
#   0.7037915 * P[status = "alive" | ulcer = "present"]
# P[status = "alive" | ulcer = "present"] = 1 - 0.7704698 * P[status = "alive" | ulcer = "present"]
# P[status = "alive" | ulcer = "present"] + 0.7704698 * P[status = "alive" | ulcer = "present"] = 1
# 1.77047 * P[status = "alive" | ulcer = "present"] = 1
# P[status = "alive" | ulcer = "present"] = 1 / 1.77047 = 0.5648218

# Now we can can calculate the  probability of dying from melanoma for a 50-year 
# old woman with tumor ulceration:
# P[status = "death from melanoma" | ulcer = "present"] = 0.7037915 * 0.5648218 = 0.3975168

odds_deathMel * (1 / (1 + odds_deathOther + odds_deathMel))

```


<!-- #.....................................................................# -->
<!-- #.....................................................................# -->


## Exercise 2

Using data from a questionnaire on the benefits of mammography (source:  Applied Logistic Regression, 2nd edition, by Hosmer and Lemeshow, 2000) completed by 412 women, we would like to investigate women's agreement to the statement: ‘You do not need a mammogram unless you develop symptoms.’, depending on their mother's and sister's breast cancer history.

```{r}

tab_SYMPT_HIST <- table(mammoexp$SYMPT, mammoexp$HIST)

mosaicplot(t(tab_SYMPT_HIST), 
           xlab = "Mother or sister with breast cancer",
           ylab = "No need for mammogram before symptoms", main = "")


tab_SYMPT_HIST <- rbind(tab_SYMPT_HIST, colSums(tab_SYMPT_HIST))
tab_SYMPT_HIST <- cbind(tab_SYMPT_HIST, rowSums(tab_SYMPT_HIST))


row.names(tab_SYMPT_HIST) <- c(
  row.names(tab_SYMPT_HIST)[-nrow(tab_SYMPT_HIST)], "total")

colnames(tab_SYMPT_HIST) <- c(
  colnames(tab_SYMPT_HIST)[-ncol(tab_SYMPT_HIST)], "total")

tab_SYMPT_HIST

```


<!-- #.....................................................................# -->

#### (a) Univariable ordinal logistic regression model
Fit an ordinal logistic regression model with `SYMPT` as dependent variable and `HIST` as explanatory variable. Which coefficient shows the change in the log-odds for women with a mother and/or sister with breast cancer history)? Is there evidence for an association between women's agreement to the statement and their mother's and sister's breast cancer history?

```{r, message = FALSE}

ordLog_mammo <- polr(SYMPT ~ HIST, data = mammoexp, method = "logistic")

ordLog_mammo_sum <- summary(ordLog_mammo)
ordLog_mammo_sum


confint(ordLog_mammo)

```


<!-- #.....................................................................# -->

#### (b) Odds estimated from the univariable ordinal logistic model
What's the odds of strongly agreeing or agreeing versus disagreeing or strongly disagreeing to the statement of interest for women with a mother and/or sister with breast cancer history, and for women whose mother and sister didn't have breast cancer history? What do the other intercepts mean?

```{r}

exp(ordLog_mammo_sum$coefficients["Agree|Disagree", "Value"] + 
      ordLog_mammo_sum$coefficients["HISTYes", "Value"]) 

exp(ordLog_mammo_sum$coefficients["Agree|Disagree", "Value"]) 

```


<!-- #.....................................................................# -->

#### (c) Proportional odds assumption meaning and visualization
The ordinal logistic model we fitted in (a) assumes proportionality of odds. What does this mean? Plot the different odds estimated by the ordinal logistic model.

```{r}

plot(x = 1, ylim = c(0, 4), xlim = c(0, 1), type = "n", xaxs = "i", xaxt = "n",
     yaxt = "n", yaxs = "i", ylab = "odds", 
     xlab = "mother and/or sister with breast cancer history")
axis(side = 1, at = c(0, 1), labels = c("No", "Yes"))
axis(side = 2, at = c(0, round(exp(ordLog_mammo_sum$coefficients[
  which(row.names(ordLog_mammo_sum$coefficients) != "HISTYes"), "Value"]), 3)), 
  las = 1, cex.axis = 0.65)

abline(coef = c(
  exp(ordLog_mammo_sum$coefficients["Strongly Agree|Agree", "Value"]),
  exp(ordLog_mammo_sum$coefficients["HISTYes", "Value"])), col = "red")

abline(coef = c(
  exp(ordLog_mammo_sum$coefficients["Agree|Disagree", "Value"]),
  exp(ordLog_mammo_sum$coefficients["HISTYes", "Value"])), col = "blue")

abline(coef = c(
  exp(ordLog_mammo_sum$coefficients["Disagree|Strongly Disagree", "Value"]),
  exp(ordLog_mammo_sum$coefficients["HISTYes", "Value"])), col = "green")

legend("topleft", col = c("red", "blue", "green"), lty = rep(1, 3), legend = c(
  "Strongly Agree|Agree", "Agree|Disagree", "Disagree|Strongly Disagree"),
  cex = 0.8)

```


<!-- #.....................................................................# -->

#### (d) Test for proportional odds assumption
After fitting a multinomial model to the data (again with `SYMPT` as dependent variable and `HIST` as explanatory variable), run a test to assess if there is evidence for a violation of the proportional odds assumption. 

```{r, message = FALSE}

multinom_mammo <- multinom(SYMPT ~ HIST, data = mammoexp)

multinom_mammo_sum <- summary(multinom_mammo)

G <- -2 * ((-ordLog_mammo$deviance/2) - (-multinom_mammo$deviance/2))

1 - pchisq(q = G, df = multinom_mammo$edf - ordLog_mammo$edf)


```


<!-- #.....................................................................# -->
<!-- #.....................................................................# -->


## Exercise 3
For this exercise we use data from a matched case-control study that investigated the role of abortions in the aetiology of secondary infertility, including 100 women admitted to the University of Athens Medical School. For every patient, two healthy control subjects from the same hospital with matching age, parity, and level of education were found.

#### (a) Conditional logistic regression model
Fit a conditional logistic regression model including the number of prior induced and prior spontaneous abortions as explanatory variables. What is the impact of abortions on the risk of secondary infertility in women?

```{r}

condLog_infer <- clogit(case ~ spontaneous + induced + strata(stratum), 
                        data = infert)

summary(condLog_infer)

```




